---
title: "Use ASH with MAD adjusted se"
author: David Gerard
output: rmarkdown::html_document
date: "`r Sys.Date()`"
---

# Abstract
CATE uses a sort of variance inflation when it "calibrates" its t-statistics. They center by the median of the t-statistics and scale by the median absolute deviation (MAD). This has the effect of making them conservative. What if we scale the variances by this MAD before applying ASH? Does it improve calibration? The answer is always yes.

# Analysis

Source poisson-thinning datamaker.
```{r}
source("../succotash_sims/code/datamaker_only_counts.R")
```

Set data generation conditions
```{r}
## these do not change
args_val              <- list()
args_val$log2foldsd   <- 1
args_val$tissue       <- "muscle"
args_val$path         <- "../../data/gtex_tissue_gene_reads/"
args_val$Ngene        <- 1000
args_val$log2foldmean <- 0
args_val$skip_gene    <- 0
args_val$Nsamp <- 10 ## size of group, sample size is 2 times this
args_val$nullpi <- 0.8
args_val$poisthin <- TRUE
ncontrols <- 100
```

Generate data with `r args_val$Nsamp * 2` samples, `r args_val$Ngene` genes, and `r ncontrols` control genes. The signal is normal with mean 0 and standard deviation `r args_val$log2foldsd`. The proportion of null genes is `r args_val$nullpi`.


```{r}
set.seed(14)
dout <- datamaker_counts_only(args = args_val)
which_null <- as.logical(dout$meta$null)

ctl <- rep(FALSE, length = args_val$Ngene)
ctl[sample((1:args_val$Ngene)[which_null], size = ncontrols)] <- TRUE

beta_true <- rep(0, length = args_val$Ngene)
beta_true[!which_null] <- dout$meta$true_log2foldchange
X <- as.matrix(model.matrix(~dout$input$condition))
colnames(X) <- c("Intercept", "Treatment")
Y <- t(log2(as.matrix(dout$input$counts + 1)))
num_sv <- sva::num.sv(t(Y), mod = X, method = "be")
num_sv
```

Fit various methods:

```{r, warning = FALSE}
ruv4out    <- ruv::RUV4(Y = Y, X = X[, 2, drop = FALSE],
                        ctl = ctl, Z = X[, 1, drop = FALSE], k = num_sv)
svaout     <- sva::sva(dat = t(as.matrix(Y)), mod = X, n.sv = 6)
Xsv <- cbind(X, svaout$sv)
svout <- limma::eBayes(limma::lmFit(object = t(Y), design = Xsv))
caterr <- cate::cate(~Treatment, X.data = as.data.frame(X), Y = Y, r = num_sv)
catenc <- cate::cate(~Treatment, X.data = as.data.frame(X), Y = Y, r = num_sv,
                     nc = ctl, adj.method = "nc")
catenc_nocal <- cate::cate(~Treatment, X.data = as.data.frame(X), Y = Y, r = num_sv,
                     nc = ctl, adj.method = "nc", calibrate = FALSE)
```

Get cate se's
```{r}
sebetahat_caterr <- sqrt(caterr$beta.cov.row) * sqrt(diag(caterr$beta.cov.col)) /
    sqrt(args_val$Nsamp * 2)
sebetahat_catenc <- sqrt(catenc$beta.cov.row) * sqrt(diag(catenc$beta.cov.col)) /
    sqrt(args_val$Nsamp * 2)

## implementation check
plot(catenc_nocal$beta, catenc$beta)
abline(0, 1)
plot(catenc_nocal$beta / sebetahat_catenc, catenc_nocal$beta.t)
abline(0, 1)
```

# MAD adjustments and ASH
If the MAD inflation improves ASH, it should take alter the $\pi_0$ estimates to be closer to `r args_val$nullpi`.

## MAD adjustment for RUV4 improves ASH.
```{r, warning = FALSE}
betahat_ruv4   <- ruv4out$beta
sebetahat_ruv4 <- sqrt(ruv4out$sigma2 * ruv4out$multiplier)
mad_mult <- mad(betahat_ruv4 / sebetahat_ruv4)
mad_mult
sebetahat_ruv4_mad <- sebetahat_ruv4 * mad_mult

ash_ruv4     <- ashr::ash(betahat = betahat_ruv4, sebetahat = sebetahat_ruv4)
ash_ruv4_mad <- ashr::ash(betahat = betahat_ruv4, sebetahat = sebetahat_ruv4_mad)
ruvash       <- ashr::ash_ruv(Y = Y, X = X, ctl = ctl, k = num_sv)

ash_ruv4$fitted.g$pi[1]
ash_ruv4_mad$fitted.g$pi[1]
ruvash$fitted.g$pi[1]
```

## MAD adjustment for SVA improves ASH.
```{r}
betahat_sva   <- svout$coefficients[, 2]
sebetahat_sva <- svout$sigma * svout$stdev.unscaled[, 2]
mad_mult <- mad(betahat_sva / sebetahat_sva)
mad_mult
sebetahat_sva_mad <- sebetahat_sva * mad_mult

ash_sva     <- ashr::ash(betahat = betahat_sva, sebetahat = sebetahat_sva)
ash_sva_mad <- ashr::ash(betahat = betahat_sva, sebetahat = sebetahat_sva_mad)

ash_sva$fitted.g$pi[1]
ash_sva_mad$fitted.g$pi[1]
ruvash$fitted.g$pi[1]
```


## MAD adjustment for CATEnc improves ASH
```{r}
betahat_catenc   <- catenc$beta
sebetahat_catenc <- sebetahat_catenc
mad_mult <- mad(betahat_catenc / sebetahat_catenc)
mad_mult
sebetahat_catenc_mad <- sebetahat_catenc * mad_mult

ash_catenc     <- ashr::ash(betahat = betahat_catenc, sebetahat = sebetahat_catenc)
ash_catenc_mad <- ashr::ash(betahat = betahat_catenc, sebetahat = sebetahat_catenc_mad)

ash_catenc$fitted.g$pi[1]
ash_catenc_mad$fitted.g$pi[1]
ruvash$fitted.g$pi[1]
```

## MAD adjustment for CATErr improves ASH.
```{r}
betahat_caterr   <- caterr$beta
sebetahat_caterr <- sebetahat_caterr
mad_mult <- mad(betahat_caterr / sebetahat_caterr)
mad_mult
sebetahat_caterr_mad <- sebetahat_caterr * mad_mult

ash_caterr     <- ashr::ash(betahat = betahat_caterr, sebetahat = sebetahat_caterr)
ash_caterr_mad <- ashr::ash(betahat = betahat_caterr, sebetahat = sebetahat_caterr_mad)

ash_caterr$fitted.g$pi[1]
ash_caterr_mad$fitted.g$pi[1]
ruvash$fitted.g$pi[1]
```




```{r}
sessionInfo()
```
