---
author: "David Gerard"
title: "Bias adjustment in VICAR"
date: "`r Sys.Date()`"
output: rmarkdown::html_document
bibliography: "sva_bib.bib"
---

# Abstract

In addition to variance inflation, I perform bias correction in the second step of VICAR. I find that there is a small improvement in terms of AUC but little difference, and perhaps slightly worse performance, in estimating $\pi_0$.

# Details

When estimating the empirical null, @efron2004large and
@van2016controlling also estimate the mean of the empirical
    null to correct for bias in the test statistics. Motivated by
    this, we replace the second equation of VICAR with
    \begin{align}
      \boldsymbol{Y}_{2\mathcal{C}} &=
      \boldsymbol{\mu}\boldsymbol{1}_p^T +
      \boldsymbol{Z}_2\hat{\boldsymbol{\alpha}}_{\mathcal{C}} +
      \boldsymbol{E}_{2\mathcal{C}},
    \end{align}
    where $\boldsymbol{\mu} \in \mathbb{R}^{k_2}$ is the bias and
    $\boldsymbol{1}_p$ is a $p$-vector of $1$'s. We then estimate
    $\boldsymbol{\mu}$ along with $\boldsymbol{Z}_2$ and
    $\lambda$. However, this is equivalent to just setting
    $\tilde{\boldsymbol{\alpha}} =
    {{\boldsymbol{1}_p^T}\choose{\hat{\boldsymbol{\alpha}}}}$ and
    estimating $(\boldsymbol{\mu}, \boldsymbol{Z}_2)$ with the previously
	described variance inflation method using
    $\tilde{\boldsymbol{\alpha}}$ instead of
    $\hat{\boldsymbol{\alpha}}$

# Results

## `ashr` $\hat{\pi}_0$

```{r, echo = FALSE}
library(ggplot2)
library(reshape2)

## ASH results ----------------------------------------------------------------
pi0hat_ash <- read.csv("pi0hat_ash.csv")

names(pi0hat_ash) <- c("Seed", "Proportion Null", "Sample Size",
                       "Number Controls", "poisthin", "OLS", "RUV2",
                       "RUV4", "RUVinv", "SVA", "CATEnc", "CATErr", "VICAR",
                       "VICARb")

pi0hat_ash$`Sample Size` <- pi0hat_ash$`Sample Size` * 2

longdat <- melt(pi0hat_ash,
                id.vars = c("Seed", "Sample Size", "Proportion Null", "Number Controls"),
                measure.vars = colnames(pi0hat_ash)[-(1:5)])


longdat_null <- longdat[longdat$`Proportion Null` == 1, ]
ggplot(longdat_null, mapping = aes(x = variable, y = value, fill = variable)) +
    geom_boxplot() +
    facet_grid("Controls" + `Number Controls`~ "Sample Size" + `Sample Size`) +
    theme_bw() +
    theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 1),
          legend.position = "none",
          text = element_text(size = 20)) +
    xlab("Methods") +
    ylab("Estimated Proportion Null") +
    geom_hline(yintercept = 1, color = "red", lty = 2) +
    geom_vline(xintercept = 7.5, color = "blue", lty = 2, alpha = 1/2)

longdat_half <- longdat[longdat$`Proportion Null` == 0.9, ]
ggplot(longdat_half, mapping = aes(x = variable, y = value, fill = variable)) +
    geom_boxplot() +
    facet_grid("Controls" + `Number Controls`~ "Sample Size" + `Sample Size`) +
    theme_bw() +
    theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 1),
          legend.position = "none",
          text = element_text(size = 20)) +
    xlab("Methods") +
    ylab("Estimated Proportion Null") +
    geom_hline(yintercept = 0.9, color = "red", lty = 2) +
    geom_vline(xintercept = 7.5, color = "blue", lty = 2, alpha = 1/2)

longdat_half <- longdat[longdat$`Proportion Null` == 0.5, ]
ggplot(longdat_half, mapping = aes(x = variable, y = value, fill = variable)) +
    geom_boxplot() +
    facet_grid("Controls" + `Number Controls`~ "Sample Size" + `Sample Size`) +
    theme_bw() +
    theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 1),
          legend.position = "none",
          text = element_text(size = 20)) +
    xlab("Methods") +
    ylab("Estimated Proportion Null") +
    geom_hline(yintercept = 0.5, color = "red", lty = 2) +
    geom_vline(xintercept = 7.5, color = "blue", lty = 2, alpha = 1/2)


```

## `qvalue` $\hat{\pi}_0$

```{r, echo = FALSE}
## qvalue results ---------------------------------------------------------
rm(list = ls())
pi0hat_qv <- read.csv("pi0hat_qvalue.csv")
names(pi0hat_qv) <- c("Seed", "Proportion Null", "Sample Size",
                      "Number Controls", "poisthin", "OLS", "RUV2",
                      "RUV4", "RUVinv", "SVA", "CATEnc", "Cal CATEnc",
                      "CATErr", "Cal CATErr", "VICAR", "VICARb")

pi0hat_qv$`Sample Size` <- pi0hat_qv$`Sample Size` * 2


longdat <- melt(pi0hat_qv,
                id.vars = c("Seed", "Sample Size", "Proportion Null", "Number Controls"),
                measure.vars = colnames(pi0hat_qv)[-(1:5)])

longdat_null <- longdat[longdat$`Proportion Null` == 1, ]
ggplot(longdat_null, mapping = aes(x = variable, y = value, fill = variable)) +
    geom_boxplot() +
    facet_grid("Controls" + `Number Controls`~ "Sample Size" + `Sample Size`) +
    theme_bw() +
    theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 1),
          legend.position = "none",
          text = element_text(size = 20)) +
    xlab("Methods") +
    ylab("Estimated Proportion Null") +
    geom_hline(yintercept = 1, color = "red", lty = 2) +
    geom_vline(xintercept = 9.5, color = "blue", lty = 2, alpha = 1/2)


longdat_half <- longdat[longdat$`Proportion Null` == 0.9, ]
ggplot(longdat_half, mapping = aes(x = variable, y = value, fill = variable)) +
    geom_boxplot() +
    facet_grid("Controls" + `Number Controls`~ "Sample Size" + `Sample Size`) +
    theme_bw() +
    theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 1),
          legend.position = "none",
          text = element_text(size = 20)) +
    xlab("Methods") +
    ylab("Estimated Proportion Null") +
    geom_hline(yintercept = 0.9, color = "red", lty = 2) +
    geom_vline(xintercept = 9.5, color = "blue", lty = 2, alpha = 1/2)

longdat_half <- longdat[longdat$`Proportion Null` == 0.5, ]
ggplot(longdat_half, mapping = aes(x = variable, y = value, fill = variable)) +
    geom_boxplot() +
    facet_grid("Controls" + `Number Controls`~ "Sample Size" + `Sample Size`) +
    theme_bw() +
    theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 1),
          legend.position = "none",
          text = element_text(size = 20)) +
    xlab("Methods") +
    ylab("Estimated Proportion Null") +
    geom_hline(yintercept = 0.5, color = "red", lty = 2) +
    geom_vline(xintercept = 9.5, color = "blue", lty = 2, alpha = 1/2)
```

## `ashr` AUC

```{r, echo = FALSE}
## ASH AUC ----------------------------------------------------------------------
rm(list = ls())
auc_ash <- read.csv("auc_ash.csv")

names(auc_ash) <- c("Seed", "Proportion Null", "Sample Size",
                       "Number Controls", "poisthin", "OLS", "RUV2",
                       "RUV4", "RUVinv", "SVA", "CATEnc", "CATErr", "VICAR",
                       "VICARb")

auc_ash$`Sample Size` <- auc_ash$`Sample Size` * 2

longdat <- melt(auc_ash,
                id.vars = c("Seed", "Sample Size", "Proportion Null", "Number Controls"),
                measure.vars = colnames(auc_ash)[-(1:5)])


longdat_half <- longdat[longdat$`Proportion Null` == 0.9, ]
ggplot(longdat_half, mapping = aes(x = variable, y = value, fill = variable)) +
    geom_boxplot() +
    facet_grid("Controls" + `Number Controls`~ "Sample Size" + `Sample Size`) +
    theme_bw() +
    theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 1),
          legend.position = "none",
          text = element_text(size = 20)) +
    xlab("Methods") +
    ylab("AUC") +
    geom_vline(xintercept = 7.5, color = "blue", lty = 2, alpha = 1/2)

longdat_half <- longdat[longdat$`Proportion Null` == 0.5, ]
ggplot(longdat_half, mapping = aes(x = variable, y = value, fill = variable)) +
    geom_boxplot() +
    facet_grid("Controls" + `Number Controls`~ "Sample Size" + `Sample Size`) +
    theme_bw() +
    theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 1),
          legend.position = "none",
          text = element_text(size = 20)) +
    xlab("Methods") +
    ylab("AUC") +
    geom_vline(xintercept = 7.5, color = "blue", lty = 2, alpha = 1/2)
```

## pvalues AUC

```{r, echo = FALSE}
## qvalue auc ---------------------------------------------------------
rm(list = ls())
auc_pvalues <- read.csv("auc_pvalues.csv")
names(auc_pvalues) <- c("Seed", "Proportion Null", "Sample Size",
                      "Number Controls", "poisthin", "OLS", "RUV2",
                      "RUV4", "RUVinv", "SVA", "CATEnc", "Cal CATEnc",
                      "CATErr", "Cal CATErr", "VICAR", "VICARb")

auc_pvalues$`Sample Size` <- auc_pvalues$`Sample Size` * 2


longdat <- melt(auc_pvalues,
                id.vars = c("Seed", "Sample Size", "Proportion Null", "Number Controls"),
                measure.vars = colnames(auc_pvalues)[-(1:5)])

longdat_half <- longdat[longdat$`Proportion Null` == 0.9, ]
ggplot(longdat_half, mapping = aes(x = variable, y = value, fill = variable)) +
    geom_boxplot() +
    facet_grid("Controls" + `Number Controls`~ "Sample Size" + `Sample Size`) +
    theme_bw() +
    theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 1),
          legend.position = "none",
          text = element_text(size = 20)) +
    xlab("Methods") +
    ylab("AUC") +
    geom_vline(xintercept = 9.5, color = "blue", lty = 2, alpha = 1/2)

longdat_half <- longdat[longdat$`Proportion Null` == 0.5, ]
ggplot(longdat_half, mapping = aes(x = variable, y = value, fill = variable)) +
    geom_boxplot() +
    facet_grid("Controls" + `Number Controls`~ "Sample Size" + `Sample Size`) +
    theme_bw() +
    theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 1),
          legend.position = "none",
          text = element_text(size = 20)) +
    xlab("Methods") +
    ylab("AUC") +
    geom_vline(xintercept = 9.5, color = "blue", lty = 2, alpha = 1/2)
```



```{r}
sessionInfo()
```

# References
