---
title: "Check if Posterior is Approximately Multivariate Normal"
author: "David Gerard"
date: "`r Sys.Date()`"
output: rmarkdown::html_document
---

## Introduction

Here, I look to see if normality is a good approximation. Also, I'm
going to see if the individual $\beta$'s can be assumed to be
independent.

Both of these conditions occur asymptotically in CATE, so it would
make sense if this asymptotic behavior also holds for RUVB. But what
about a finite number of samples?

In this write-up, I find that the tails are a little too big to be
normal when the sample size is small. The normality assumption looks
great for larger sample sizes. The independence assumption seems
reasonable.

## Bladder Data

```{r}
library(ggplot2)
bladder <- readRDS(file = "./output/ruvbout_q75/ruvbout_bladder.Rds")
beta    <- bladder$betahat_post[1,,]
```

I read in the RUVB output when using the GTEx bladder data using
gender as the covariate of interest. I included an intercept term and
the 75th percentile of the $\log_2$ counts as nuisance covariates. I
am choosing to focus on the bladder data because with n = 11, it has
the smallest sample size of the tissues in the GTEx data.

## Look at correlations

For the `r nrow(beta)` by `r nrow(beta)` correlation matrix of the
$\beta$'s, the maximum correlation observed is 0.1, and the
correlation is unimodal about 0.

```{r}
cb <- cor(t(beta[sample(1:nrow(beta), size = 1000), ]))
qplot(cb[lower.tri(cb)], bins = 30)
```

But if we have independent normals, we would expect to see some
heterogeneity between the correlations. Let's look at a histogram from
theoretically independent normals.

```{r}
temp <- matrix(rnorm(1000 ^ 2), nrow = 1000, ncol = 1000)
ct <- cor(temp)
qplot(ct[lower.tri(ct)], bins = 30)
```

It looks the exact same. A qqplot of the correlations we get

```{r}
temp <- data.frame(sample = sort(cb[lower.tri(cb)]), theoretical = sort(ct[lower.tri(ct)]))
ggplot(data = temp, mapping = aes(x = theoretical, y = sample)) + geom_point() +
    geom_smooth(method = "lm", lty = 2)
```


## Look at tail behavior

The tails in the bladder data are always a little too big. Here are some univariate qqplots:

```{r}
plot_random_qq <- function(beta, seed) {
    set.seed(seed)
    temp <- data.frame(sample = beta[sample(1:nrow(beta), 1), ])
    temp$theoretical <- qnorm(seq(1 / nrow(temp), 1 - 1 / nrow(temp),
                           length = nrow(temp)))[rank(temp$sample)]
    pl <- ggplot(data = temp, mapping = aes(x = theoretical, y = sample)) + geom_point() +
        geom_smooth(method = "lm", lty = 2)
    print(pl)
}
for (index in 1:10 + 301) {
    plot_random_qq(beta, index)
}
```

Here are some random scatterplots

```{r, eval = FALSE, echo = FALSE}
plot_random_scatter <- function(beta, seed) {
    set.seed(seed)
    temp <- data.frame(b1 = beta[sample(1:nrow(beta), 1), ], b2 = beta[sample(1:nrow(beta), 1), ])
    pl <- ggplot(data = temp, mapping = aes(x = b1, y = b2)) + geom_point()
    print(pl)
}
for (index in 1:10 + 5000) {
    plot_random_scatter(beta, index)
}
```

## Muscle Data
The muscle data has a much larger sample size (n = 430). The qqplots look great.
```{r}
muscle <- readRDS(file = "./output/ruvbout_q75/ruvbout_muscle.Rds")
beta   <- muscle$betahat_post[1,,]
for (index in 1:10 + 5325) {
    plot_random_qq(beta, index)
}
```
