---
title: "Initial GTEX Analysis"
author: David Gerard
output: pdf_document
date: "`r Sys.Date()`"
bibliography: gtex_bib.bib
---

# Abstract
For small n, in all null setting, example of how other methods have poor performance.

# Brief Summary
Read in the cleaned heart gtex data and estimate the number of hidden confounders with `sva::num.sv()`.
```{r, cache = TRUE}
set.seed(816)
library(ggplot2)
library(xtable)
load(file = "../output/cleaned_gtex_data/heart.Rd")
n <- 20
Y <- heart$Y[, sample(1:ncol(heart$Y), size = n)]
X <- model.matrix(~sample(c(0, 1), size = ncol(Y), replace = TRUE))
colnames(X) <- c("Intercept", "Treatment")
ctl <- heart$ctl
k_large <- sva::num.sv(dat = heart$Y, mod = heart$X)
k <- sva::num.sv(dat = Y, mod = X)
k
```

$Y$ is log(COUNTS + 1) of the top `r nrow(heart$Y)` expressed genes (genes with at least 10 reads per sample on average). Samples from the same individual were averaged, resulting in $n =$ `r ncol(heart$Y)`. $X$ contains an intercept and a random indicator for "treatment" vs "control". Since the treatment label is random, all genes are null.

Control genes are the housekeeping genes identified by @eisenberg2013human. This is a different list than that used by @gagnon2012using, which was drawn from @eisenberg2003human. In total, there are `r sum(heart$ctl)` housekeeping genes among the `r nrow(heart$Y)` most expressed genes in the GTEX heart data.

# Analysis

## Fit RUVASH
When I fit RUVASH on the small data, it works great
```{r, cache = TRUE}
ruvash_out <- ashr::ash_ruv(Y = t(Y), X = X, ctl = ctl, k = k,
                            posthoc_inflate = TRUE)
ruvash_out$fitted.g$pi[1]
qvalue::qvalue(ruvash_out$ruv$pvalues)$pi0

hist(ruvash_out$ruv$pvalues, main = "RUVASH p-values")
hist(ruvash_out$ruv$pvalues[ctl], main = "RUVASH p-values of control genes")
```

## Fit competitors
OLS
```{r, cache = TRUE}
limmaout   <- limma::lmFit(Y, X)
betahat_ols    <- limmaout$coefficients[, 2]
sebetahat_ols  <- limmaout$stdev.unscaled[, 2] * limmaout$sigma
ols_p      <- 2 * (1 - pt(abs(betahat_ols / sebetahat_ols), df = nrow(X) - ncol(X)))
ols_q      <- qvalue::qvalue(ols_p)
ols_a      <- ashr::ash(betahat = betahat_ols, sebetahat = sebetahat_ols)

ols_q$pi0
ols_a$fitted.g$pi[1]

hist(ols_p, main = "OLS p-values")
hist(ols_p[ctl], main = "OLS p-values for control genes")
```

SVA
```{r, cache = TRUE}
trash      <- capture.output(sva_out <- sva::sva(dat = Y, mod = X, n.sv = k))
X.sv       <- cbind(X, sva_out$sv)
limmaout   <- limma::lmFit(Y, X.sv)
betahat_sva    <- limmaout$coefficients[, 2]
sebetahat_sva  <- limmaout$stdev.unscaled[, 2] * limmaout$sigma
sva_p      <- 2 * (1 - pt(abs(betahat_sva / sebetahat_sva), df = nrow(X.sv) - ncol(X.sv)))
sva_q      <- qvalue::qvalue(sva_p)
sva_a      <- ashr::ash(betahat = betahat_sva, sebetahat = sebetahat_sva)

sva_q$pi0
sva_a$fitted.g$pi[1]

hist(sva_p, main = "SVA p-values")
hist(sva_p[ctl], main = "SVA p-values for control genes")
```
RUV2 and RUV4
```{r, cache = TRUE}
ruv4out <- ruv::RUV4(Y = t(Y), X = X[, 2, drop = FALSE], ctl = ctl,
                     Z = X[, -2, drop = FALSE], k = k)
betahat   <- ruv4out$betahat
sebetahat <- sqrt(ruv4out$sigma2 * ruv4out$multiplier)
ruv4_q    <- qvalue::qvalue(ruv4out$p)
ruv4_a    <- ashr::ash(betahat = betahat, sebetahat = sebetahat)
ruv4_q$pi0
ruv4_a$fitted.g$pi[1]

hist(ruv4out$p, main = "RUV4 p-values")
hist(ruv4out$p[ctl], main = "RUV4 p-values for control genes")


ruv2out <- ruv::RUV2(Y = t(Y), X = X[, 2, drop = FALSE], ctl = ctl,
                     Z = X[, -2, drop = FALSE], k = k)
betahat   <- ruv2out$betahat
sebetahat <- sqrt(ruv2out$sigma2 * ruv2out$multiplier)
ruv2_q    <- qvalue::qvalue(ruv2out$p)
ruv2_a    <- ashr::ash(betahat = betahat, sebetahat = sebetahat)
hist(ruv2out$p, main = "RUV2 p-values")
hist(ruv2out$p[ctl], main = "RUV2 p-values for control genes")

ruv2_q$pi0
ruv2_a$fitted.g$pi[1]
```
CATErr and CATEnc
```{r, cache = TRUE}
cate_nc <- cate::cate(~Treatment, Y = t(Y),
                      X.data = as.data.frame(X[, 2, drop = FALSE]),
                      r = k, adj.method = "nc", fa.method = "pc",
                      nc = ctl, calibrate = FALSE)
betahat_catenc   <- cate_nc$beta
sebetahat_catenc <- sqrt(cate_nc$beta.cov.row * cate_nc$beta.cov.col) / sqrt(nrow(X))
catenc_q  <- qvalue::qvalue(cate_nc$beta.p.value)
catenc_a  <- ashr::ash(betahat_catenc, sebetahat_catenc)
catenc_q$pi0
catenc_a$fitted.g$pi[1]

hist(cate_nc$beta.p.value, main = "CATE NC p-values")
hist(cate_nc$beta.p.value[ctl], main = "CATE NC p-values for control genes")
```

```{r, cache = TRUE}
cate_rr <- cate::cate(~Treatment, Y = t(Y),
                      X.data = as.data.frame(X[, 2, drop = FALSE]),
                      r = k, adj.method = "rr", fa.method = "pc",
                      calibrate = FALSE)
betahat   <- cate_rr$beta
sebetahat <- sqrt(cate_rr$beta.cov.row * cate_rr$beta.cov.col) / sqrt(nrow(X))
caterr_q  <- qvalue::qvalue(cate_rr$beta.p.value)
caterr_a  <- ashr::ash(betahat, sebetahat)
caterr_q$pi0
caterr_a$fitted.g$pi[1]

hist(cate_rr$beta.p.value, main = "CATErr p-values")
hist(cate_rr$beta.p.value[ctl], main = "CATErr p-values for control genes")
```

## Compare $\pi_0$ estimates


## Calibrated CATE
```{r, cache = TRUE}
cate_nc_cal <- cate::cate(~Treatment, Y = t(Y),
                      X.data = as.data.frame(X[, 2, drop = FALSE]),
                      r = k, adj.method = "nc", fa.method = "pc",
                      nc = ctl, calibrate = TRUE)
hist(cate_nc_cal$beta.p.value, main = "CATE NC Calibrated p-values")
hist(cate_nc_cal$beta.p.value[ctl], main = "CATE NC Calibrated p-values for control genes")
cate_nc_cal_q <- qvalue::qvalue(cate_nc_cal$beta.p.value)
cate_nc_cal_q$pi0

cate_rr_cal <- cate::cate(~Treatment, Y = t(Y),
                      X.data = as.data.frame(X[, 2, drop = FALSE]),
                      r = k, adj.method = "rr", fa.method = "pc",
                      calibrate = TRUE)
hist(cate_rr_cal$beta.p.value, main = "CATErr Calibrated p-values")
hist(cate_rr_cal$beta.p.value[ctl], main = "CATErr Calibrated p-values for control genes")
cate_rr_cal_q <- qvalue::qvalue(cate_rr_cal$beta.p.value)
cate_rr_cal_q$pi0
```

```{r}
sessionInfo()
```

# References
